---
title: "class09: Candy Mini Project"
author: "Ethan Arevalo (A17393064)"
format: pdf
toc: TRUE
---

## Background

In this mini-project, you will explore FiveThirtyEight's Halloween candy set. FiveThirtyEight, sometimes rendered as just 538, is an American website that focuses mostly on opinion poll analysis, politics, economics, and sports blogging. We will **ggplot** some basic stats, correlation analysis and PCA to make sense of the landscape of US candy

transcriptomics work that we will use these methods on throughout the rest of the course

## Data Import

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
There are `r nrow(candy)` rows in this dataset

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
library(dplyr)

candy |> filter((row.names(candy)) == "3 Musketeers") |>
select(winpercent)
```

```{r}
candy["3 Musketeers", "winpercent"]
```


> Q4. What is the winpercent value for “Kit Kat”?


```{r}
candy["Kit Kat", "winpercent"]
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", "winpercent"]
```


> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes! "Winpercent"

> Q7. What do you think a zero and one represent for the `candy$chocolate` column?

```{r}
candy$chocolate
```

A 1 would 

## Exploratory analysis

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```
> Q9. Is the distribution of winpercent values symmetrical?

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins = 15)
```

The distribution is not symmetrical 

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.candy <- candy[candy$chocolate == 1, ]
choc.win <- choc.candy$winpercent
mean(choc.win)
```

```{r}
fruit.candy <- candy[candy$fruity == 1, ]
fruit.win <- fruit.candy$winpercent
mean(fruit.win)
```

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```

##Overall candy rankings

> Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)

candy |> arrange(winpercent) |> head(5)
```

```{r}
ord.id <- order(candy$winpercent)
candy[ord.id, ]
```


> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |> arrange(desc(winpercent)) |> head(5)
```

```{r}
tail(candy[ord.id, ])
```


> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```
> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?


```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col() +
  ylab("")
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburts

```{r}
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(x=winpercent, y=pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 17)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$winpercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Chocolate has the highest winpercent with a variety of low prices 

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

Nik L Nip is the least popular of the most expensive candies

## Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate

> Q23. Similarly, what two variables are most positively correlated?

Any variable against itself 

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (chocolate), chocolate other (light brown), fruity (pink), other (black)",
       caption="Data from 538")
```

```{r}
#library(plotly)
#ggplotly(p)
```

```{r}
ggplot(pca$rotation) +
  aes(x=PC1, y=reorder(rownames(pca$rotation), PC1)) +
  geom_col()
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

Fruity, pluribus, and hard. This makes sense when referring to the Halloween candy OCA in which fruity had a positive correlation 

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

Fruity, pluribus and hard
