---
title: "Class 08: Breast Cancer Mini Project"
author: "Ethan Arevalo (A17393064)"
format: pdf
toc: TRUE
---

## Background

In todays class we will apply the methods and techniques clustering and PCA to help make sense of a real world breast cancer FNA biopsy data set. 

## Data Import

We start by importing our data. It is a CSV file so we will use the `read.csv()` function

```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
head(wisc.df)
```

Make sure to remove the first `diagnosis` column - I don't want to use this for my machine learning models. We will use it later to compare our results to the expert diagnosis

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(wisc.df$diagnosis == "M")
```

```{r}
table(wisc.df$diagnosis)
```


> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
colnames(wisc.data)
length(grep("_mean", colnames(wisc.data)))
```


```{r}
ncol(wisc.data)
```

##Principal Component Analysis

The main function here is `prcomp()` and we want to make sure we set the optional argument `scale=TRUE`:

```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```


```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

0.427

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

PC3

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

PC7

```{r}
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

All the red lines lead to one point on the graph but due to all the data and its composition, it is very hard to understand

Our main PCA score plot or "PC plot" of results:

```{r}
library(ggplot2)
```
```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2, col = diagnosis) +
  geom_point()
```

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col = diagnosis) +
  geom_point()
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

PC1 vs PC2 has a wider range with PC2 going past -10 on the y axis whereas, PC1 vs PC3 is more concentrated 

Collectively these two plot ("score plot" and "loading plot") tell us that if cells nucleus are deeply indented ("concave"), irregular non circular ("compactness"), and have larger "perimeter" values they tend to be malignant...

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

## Hierarchical clustering 

First scale the data `scale()` function, then calculate a distance matrix (`dist()`). Then cluster with the `hclust()` function and plot:

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
```

```{r}
data.dist <- dist(data.scaled)
```

```{r}
wisc.hclust <- hclust(data.dist)
```

> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=20, col="red")
```

You can also use `cutree()` function with an argument `k=4`rather than `h=height`

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters)
```

## Combining Methods

Here we will take our PCA results and use those as input for clustering. In other words our `wisc.pr$x` scores that we plotted above (the main output from PCA - how the data lie on our new principle component axis/variables) and use a subset of the PCs as input for `hclust()`

```{r}
pc.dist <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```

Cut the dendrogram/tree into two main groups/clusters:

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

I want to know how the clustering in `grps` with the values of 1 or 2 correspond the expert `diagnosis`

```{r}
table(grps, diagnosis)
```

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(col=grps)
```

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```

```{r}
table(wisc.hclust.clusters, diagnosis)
```


My clustering **groups 1** are mostly "M" diagnosis (179) and my clustering **group 2** are mostly "B" diagnosis 

24 FP
179 TP
TN 333
FN 33

Sensitivity TP/(TP+FN)
```{r}
179/(179+33)
```

Specificity TN/(TN+FP)
```{r}
333/(333+24)
```

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

