---
title: "class13: RNASeq with DESeq2"
author: "Ethan Arevalo (A17393064)"
format: pdf
toc: TRUE
---

## Background
Today we will perform an RNASeq analysis of the effects of a common steroid on airway cells 

In paticular, dexamethasone (herafter just called "dex") on different airway smooth muscle cell lines (ASM cells)

## Data Import

We need two different inputs:

- **countData**: with genes in rows and experiments in columns 
- **colData**: meta data that describes the columns in countData

```{r}
library("DESeq2")
```

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <- read.csv("airway_metadata.csv")
```

```{r}
head(counts)
```

```{r}
head(metadata)
```

> Q1. How many genes are in this dataset

```{r}
nrow(counts)
```

> Q2. How many control cell lines do we have 

```{r}
table(metadata$dex)
```

## Differential gene expression

We have 4 replicate drug treated and control (no drug) columns/experiments in our `counts` object

We want one "mean" value for each gene (rows) in "treated" (drug) and one mean value for each gene in "control" cols. 

> Q3. 

Step1. Find all "control" columns
```{r}
control.inds <- metadata$dex == "control"
```

Step2. Extract these columns to a new object called `control.counts` 
```{r}
control.counts <- counts[, control.inds]
```

Step3. Then find the mean value for each gene
```{r}
control.mean <- rowMeans(control.counts)
```

> Q4. 

Now we do the same thing for the "treated" columns/experiments...

```{r}
treated.inds <- metadata$dex == "treated"
```

```{r}
treated.counts <- counts[, treated.inds]
```

```{r}
treated.mean <- rowMeans(treated.counts)
```

```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

>Q5.

A quick plot

```{r}
plot(meancounts[,1],meancounts[,2], xlab="Control", ylab="Treated")
```

>Q6.

Lets log transform this count data:

```{r}
plot(meancounts, log="xy")
```

**N.B.** We most often use log2 for this type of data as it makes the interpretation much more straightforward

Treated/Control is often called "fold-changed" 

If there was no change we would have a log2-fc of 0
```{r}
log2(10/10)
```

If we had double the amount of transcript around
```{r}
log2(20/10)
```

If we have half as much transcript around we would have a log2-fc of -1
```{r}
log2(5/10)
```

```{r}
log2(40/10)
```

> Q. Calculate a log2 fold change value for all our genes and add it as a new column to our `meancounts` object.

```{r}
meancounts$log2f <- log2(meancounts$treated.mean / 
                          meancounts$control.mean)

head(meancounts)
```

>Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?

arr.ind tells us the zero counts of both genes and samples

There are some "funky" log2fc values (NaN and -Inf) here that came about when ever we have 0 mean count values. Typically we would remove these genes from any further analysis - as we can't say anything about them if we have no data for them

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```


```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

```{r}
up.ind
down.ind
```

>Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level? 

250

>Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level? 

367

>Q10. Do you trust these results? Why or why not?

No because there is signifier that the values are statisitically different 

## DESeq analysis

Lets do this analysis with an estimate of statistical significance using the **DESeq2** package

```{r, message=FALSE}
library(DESeq2)
```

DESeq (like many bioconductor packages) want its input data in a very specific way. 

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts,
                       colData = metadata,
                       design = ~dex)
```

### Run the DESeq analysis pipeline

The main function `DESeq()`
```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
head(res)
```

```{r}
36000 * 0.05
```

## Volcano Plot

This is a main summary results figure from these kinds of studies. It is a plot of Log2 fold-change vs (Adjusted) P-value 

```{r}
plot(res$log2FoldChange,
     res$padj)
```

Again this y-axis highly needs log transforming and we can flip the y-axis with a minus sign so it looks like every other volcano plot
```{r}
plot(res$log2FoldChange,
     -log(res$padj))
abline(v=-2, col="red")
abline(v=+2, col="red")
abline(h=-log(0.05), col="red")
```

### Adding some color annotation

```{r}
mycols <- rep("gray", nrow(res))
mycols[res$log2FoldChange > 2] <- "blue"
mycols[res$log2FoldChange < -2] <- "darkgreen"
mycols[res$padj >= 0.05] <- "gray"

plot(res$log2FoldChange,
     -log(res$padj),
     col=mycols)

abline(v=c(-2, +2), lty=2)
abline(h=-log(0.05), lty=2)
```

> Q. Make a presentation quality ggplot version of this plot. Include clear axis labels, a clean theme, your custom colors, cut off lines and a plot title

```{r}
library(ggplot2)

ggplot(res) +
  aes(log2FoldChange,
      -log(padj)) +
  geom_point(col=mycols)
  labs(x="Log2 Fold-change",
     y="-log Adjusted P-value") +
  geom_vline(xintercept = c(-2, 2)) +
  geom_hline(yintercept = -log(0.05)) +
  theme_bw()
```

## Save our results

Write a csv file
```{r}
write.csv(res, file="results.csv")
```

